How to make an abridged life table and calculate life expectancy

### 2016-09-16 working_teaching_examples/how_to_abridged_life_table.Rmd
added life expectancy examples

##What is a life table?
"A life table is a system way of keeping track of the mortality experience of a group." [@selvin_text, Chapter 9 Life Tables: An Introduction].

@@

This data was used in a previously published paper [@auger2014mortality] and we use it here to show how to create an abridged life table and calculate life expectancy at birth.

@@

I then add an additional column that denotes how many years are in each age bracket:
```{r add_n_x}
canada.men$n_x = 5
canada.men$n_x[canada.men$ages == 0] = 1
canada.men$n_x[canada.men$ages == 1] = 4
```

@@

>>>>

##What is a life table?
"A life table is a systematic way of keeping track of the mortality experience of a group." [@selvin_text, Chapter 9 Life Tables: An Introduction].

@@

This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.

@@

I then add a column that denotes how many years are in each age bracket:
```{r add_n_x}
canada.men$n_x = 5
canada.men$n_x[canada.men$ages == 0] = 1
canada.men$n_x[canada.men$ages == 1] = 4
```

```{r function_lifeexpectancy}
#dataset needs to be in order of age
life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA, calendar.years = NA){

  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates

  if(is.na(ave.prop.lived) == T) {
      data["a_x"] <- 0.5 #average proportion of interval lived
      data[1, "a_x"] <- 0.09
  } else{
      data["a_x"] <- data[ave.prop.lived]
  }

  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
  data["p_x"] <- 1 - data["q_x"] #probability of survival

  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
  data[1, "l_x"] <- 100000

  for(i in 2:dim(data)[1]) {
    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
  }

  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
  data["L_x"] <- (data["l_x"] - data["d_x"])*data[num.ages.in.group] + data["d_x"]*data["a_x"]*data[num.ages.in.group] #person time in the age bracket
  data[dim(data)[1], "L_x"] <- data[dim(data)[1], "l_x"]/data[dim(data)[1], "R_x"] #different for last age bracket

  #now to calculate the total person time lived at or after age x, T_x:
  data["cumsum"] <- cumsum(data["L_x"])
  data["l.cumsum"] <- c(0, data[1:(dim(data)[1] - 1), "cumsum"])
  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
  data["e_x"] <- data["T_x"]/data["l_x"]

  return(data)

} 
```

@@

**A second example**

In this example we use the data that Sam entered into Stata to illustrate the life expectancy concept to me. I'll replicate the life table here to ensure that I can arrive at the same results as he does in Stata.

```{r lifetable_cali_abridged}
cali.abridged <- data.frame("ages" = c("0-1y", "1-4y", "5-9y", "10-14y", "15+y"), "nx" = c(1, 4, 5, 5, 5),
                         "obs.deaths" = c(1635, 168, 251, 371, 23300), #last number is fictitious
                         "obs.popn" = c(123342, 435167, 578964, 649693, 143600))

cali.abridged

life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "obs.deaths", population.counts = "obs.popn")
```

```{r lifetable_cali_complete}
cali.complete <- data.frame("ages" = c(0:15), "nx" = rep(1,16),
                         "obs.deaths" = c(1635, 64, 41, 22, 41, 55, 42, 58, 44, 52, 48, 60, 52, 82, 129, 23300), #last number is fictitious
                         "obs.popn" = c(123342, 111520, 109200, 108749, 105698, 110548, 106857, 112184, 116423,
                                        132952, 134266, 128938, 125502, 128212, 132775, 143600))

head(cali.complete)

life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "obs.deaths", population.counts = "obs.popn")

cali.complete$avelived <- 0.5
cali.complete$avelived[cali.complete$ages==0] <- 0.1

life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", ave.prop.lived = "avelived",
           death.counts = "obs.deaths", population.counts = "obs.popn")
```
